\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-05-27 doc/paper-jss/sdmTMB-paper.Rnw
Revisions

% setwd("doc/paper-jss/")
% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)

@@

%% -- Article metainformation (author, title, ...) -----------------------------

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
% TODO: decide on using GRF throughout

Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation when evaluating the multivariate normal density function.
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep{thorson2015ecology, thorson2015ices, thorson2019, osgood-zimmerman2023}.  This SPDE method approximates a Matérn correlation function as arising mechanistically from local diffusion in space and/or time \citep{lindgren2011} and it results in a sparse precision matrix that permits efficient computation using existing sparse-matrix libraries \citep{rue2005gmrf}.

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
\Rev{B2}Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.

@@

\Rev{B4a}In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{rue2005gmrf,simpson2012}.
One such approach is the SPDE approach, which approximates a mechanistic process of local diffusion using methods drawn from finite-element analysis.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develop the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} summarizes the approach and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provide a recent review of the approach and its applications over the last decade.

\Rev{B4b}For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally approximated as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma_m^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in \proglang{C++}.
% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023). _Matrix: Sparse and Dense Matrix Classes and Methods_. R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack. If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al. (2010). “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
\Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

A fitted model summary can be viewed with \fct{print} or \fct{summary} and a set of basic ``sanity'' checks can be run with \fct{sanity}.
\fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
\Rev{A4}These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.

@@

We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
\Rev{A19b}We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.

@@

<<owl-fit, echo=TRUE, results = "hide">>=
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow,
  mesh = mesh_snow,
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  reml = TRUE,
  silent = FALSE
)
@

@@

\Rev{B7}The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case, \code{zeta\_s\_nao}).
However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
We illustrate a way of accomplishing this by simulating from uncertainty about fixed and random effects while assuming that parametric uncertainty is well approximated using a multivariate normal distribution and the joint precision matrix.
We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
By default, \code{nsim > 0} will return a matrix of draws for the overall prediction.
Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the SVC random field.
This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
We then use the function \fct{spread\_sims} to draw 200 simulations for the parameters themselves.
Because the simulations are stored in different dimensions, the random field draws must be transposed \code{t()} before combining the vector of main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}.
Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
For this example, we use thresholds of 0.1 and 0.9 representing CIs of 80\%.

<<owl-p, message=FALSE, echo=FALSE, eval=FALSE, cache=TRUE>>=
# not currently including, but might be helpful
snow <- predict(fit_owl, newdata = snow)
snow |> select(X, Y, year, nao, count, est, omega_s, epsilon_st, zeta_s_nao) |>
  head(n = 2)
@

<<zeta-effect, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
dim(zeta_s)
sims <- spread_sims(fit_owl, nsim = 200)
dim(sims)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- exp(apply(combined, 2, median))
snow$nao_effect_lwr <- exp(apply(combined, 2, quantile, probs = 0.1))
snow$nao_effect_upr <- exp(apply(combined, 2, quantile, probs = 0.9))
@

@@

<<shapes, echo=FALSE>>=
if (!file.exists(here::here("data/ne_10m_lakes"))) {
  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
    "download/10m/physical/ne_10m_lakes.zip")
  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
}
@

<<shapes-read, echo=FALSE>>=
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
land <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
land <- land %>% sf::st_transform(crs = Albers)
@

<<owl-proj2>>=
# project to get nice axis limits
snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
snow2 <- snow2 |> mutate(x = X, y = Y) |>
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. The effect is multiplicative on owl count per NAO unit. In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive. Point size is scaled to the mean counts in each location.", fig.width=5, out.width="4in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
nao_effect_df <- select(snow2, X, Y, count, nao_effect, nao_effect_lwr, nao_effect_upr) |>
  group_by(X, Y) |>
  summarise_all(mean)

snow_g1 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", fill = NA, lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
    guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
  ) +
  guides(size = "none") +
  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect), 2), " to ", round(max(snow$nao_effect), 2))) +
  labs(colour = "NAO effect\non Snowy Owl counts") +
  theme_bw() +
  theme(
    legend.position = c(0.25, 0.17),
    legend.title = element_text(size = 9, hjust = 0),
    plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
    panel.background = element_rect(fill = "grey90", colour = NA),
    panel.grid.major = element_line(colour = "white", size = 0.5),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(size = 10, hjust = -0.15),
    axis.title = element_blank()
  )

>>>>

% setwd("doc/paper-jss/")
% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)
% R -e 'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)'

@@

%% -- Article meta information (author, title, ...) -----------------------------

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011, thorsonkristensen2024}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
% TODO: decide on using GRF throughout

Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation when evaluating the multivariate normal density function.
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep{thorson2015ecology, thorson2019, osgood-zimmerman2023, thorsonkristensen2024}.
This SPDE method approximates a Matérn correlation function as arising mechanistically from local diffusion in space and/or time \citep{lindgren2011} and it results in a sparse precision matrix that permits efficient computation using existing sparse-matrix libraries \citep{rue2005gmrf}.

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.

@@

In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{rue2005gmrf,simpson2012}.
One such approach is the SPDE approach, which approximates a mechanistic process of local diffusion using methods drawn from finite-element analysis.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develop the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} summarizes the approach and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provide a recent review of the approach and its applications over the last decade.

For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally approximated as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma_m^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in \proglang{C++}.
% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023). _Matrix: Sparse and Dense Matrix Classes and Methods_. R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack. If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al. (2010). “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2021}.
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

A fitted model summary can be viewed with \fct{print} or \fct{summary} and a set of basic ``sanity'' checks can be run with \fct{sanity}.
\fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Other \Rev{A7} standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
\Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.

@@

We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.

@@

<<owl-fit, echo=TRUE, results = "hide">>=
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow,
  mesh = mesh_snow,
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  silent = FALSE
)
@

@@

The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case, \code{zeta\_s\_nao}).
However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
We illustrate a way of accomplishing this by simulating from uncertainty about fixed and random effects while assuming that parametric uncertainty is well approximated using a multivariate normal distribution and the joint precision matrix.
We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
By default, \code{nsim > 0} will return a matrix of draws for the overall prediction.
Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the SVC random field.
This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
We then use the function \fct{spread\_sims} to draw 200 simulations for the parameters themselves.
Because the simulations are stored in different dimensions, the random field draws must be transposed \code{t()} before combining the vector of main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}.
Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
For this example, we use thresholds of 0.1 and 0.9 representing CIs of 80\%.

<<owl-p, message=FALSE, echo=FALSE, eval=FALSE, cache=TRUE>>=
pred_snow <- predict(fit_owl, newdata = snow)
pred_snow |>
  select(X, Y, year, nao, count, est, omega_s, epsilon_st, zeta_s_nao) |>
  head()
@

<<zeta-effect, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
set.seed(1)
zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
dim(zeta_s)
sims <- spread_sims(fit_owl, nsim = 200)
dim(sims)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- exp(apply(combined, 2, median))
snow$nao_effect_lwr <- exp(apply(combined, 2, quantile, probs = 0.1))
snow$nao_effect_upr <- exp(apply(combined, 2, quantile, probs = 0.9))
@

@@

<<shapes, echo=FALSE>>=
if (!file.exists(here::here("data/ne_10m_lakes.shp"))) {
  lakes <- rnaturalearth::ne_download(
    scale = 10, type = "lakes", category = "physical",
    destdir = here::here("data")
  )
}
@

<<shapes-read, echo=FALSE>>=
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") |>
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] |> sf::st_transform(crs = Albers)
land <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
land <- land |> sf::st_transform(crs = Albers)
@

<<owl-proj2>>=
snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
snow2 <- snow2 |> mutate(x = X, y = Y) |>
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. The effect is multiplicative on owl count per NAO unit. In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive. Point size is scaled to the mean counts in each location.", fig.width=5, out.width="4in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
nao_effect_df <- snow2 |>
  group_by(X, Y) |>
  summarise(
    count = mean(count), nao_effect = mean(nao_effect),
    nao_effect_lwr = mean(nao_effect_lwr), nao_effect_upr = mean(nao_effect_upr),
    .groups = "drop"
  )

snow_g1 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", fill = NA, lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
    guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
  ) +
  guides(size = "none") +
  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect), 2), " to ", round(max(snow$nao_effect), 2))) +
  labs(colour = "NAO effect\non Snowy Owl counts") +
  theme_bw() +
  theme(
    legend.position = c(0.25, 0.17),
    legend.title = element_text(size = 9, hjust = 0),
    plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
    panel.background = element_rect(fill = "grey90", colour = NA),
    panel.grid.major = element_line(colour = "white", linewidth = 0.5),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(size = 10, hjust = -0.15),
    axis.title = element_blank()
  )

th <- theme(
    legend.position = "none",
    panel.background = element_rect(fill = "grey90", colour = NA),
    panel.grid.major = element_line(colour = "white", size = 0.5),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(size = 10), axis.ticks = element_blank(),
    axis.title = element_blank(), axis.text = element_blank()
  )

snow_g2 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_lwr, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr), 2), " to ", round(max(snow$nao_effect_lwr), 2))) +
  theme_bw() +
  th

snow_g3 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_upr, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr), 2), " to ", round(max(snow$nao_effect_upr), 2))) +
  theme_bw() +
  th
